### 2015-03-13 background.tex
update in background section

Nowadays, multicore architectures have become the mainstream, leading to an ever-increasing demand for concurrency in operating systems. For example, modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently to perform device related operations; a running driver process can block and be replaced by another process in the same driver; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging). All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:

@@

Listing~\ref{data_race_example} is an example of a data race between two entry points of a Linux driver. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

%\begin{figure}[!h]
%\centering
%\noindent\begin{minipage}{.95\textwidth}
%\noindent\begin{minipage}{.485\textwidth}
%\begin{lstlisting}[caption={Example of a simple data race in a Linux device driver.},
%label=data_race_example]
%static void ep1 (struct shared *dev) {
%  dev->resource = 5;
%}
%
%static void ep2 (struct shared *dev) {
%  if (dev->resource != 5)
%    dev->resource = 3;
%}
%\end{lstlisting}
%\end{minipage}\hfill
%\noindent\begin{minipage}{.485\textwidth}
%\begin{lstlisting}[caption={Data race eliminated by using the highlighted locks.},
%label=lock_example]
%static void ep1 (struct shared *dev) {
%@lock(&dev->lock);@
%  dev->resource = 5;
%@unlock(&dev->lock);@
%}
%
%static void ep2 (struct shared *dev) {
%@lock(&dev->lock);@
%  if (dev->resource != 5)
%    dev->resource = 3;
%@unlock(&dev->lock);@
%}
%\end{lstlisting}
%\end{minipage}
%\end{minipage}
%\end{figure}

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Listing~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.

>>>>

Nowadays, multicore architectures have become mainstream, leading to an ever-increasing demand for concurrency in operating systems. Modern Linux kernel distributions address this demand by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same module concurrently; a running module process can block and be replaced by another process in the same module; hardware interrupts can be handled concurrently; delayed code execution is the norm; and the user can insert or remove hardware components without rebooting the kernel (known as hot-plugging). All these possibilities can easily lead to serious concurrency issues, such as \emph{data races}, which can be defined as follows:

@@

Figure~\ref{data_race_example} is an example of a data race between two entry points of a Linux module. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

\begin{figure}[!h]
\centering
\begin{lstlisting}
static void ep1 (struct shared *dev) {
  dev->resource = 5;
}

static void ep2 (struct shared *dev) {
  if (dev->resource != 5)
    dev->resource = 3;
}
\end{lstlisting}
\caption{Simple data race in a Linux module}
\label{data_race_example}
\end{figure}

\begin{figure}[!h]
\centering
\noindent\begin{minipage}{.95\textwidth}
\begin{lstlisting}
static void ep1 (struct shared *dev) {
lock(&dev->lock);
  dev->resource = 5;
unlock(&dev->lock);
}

static void ep2 (struct shared *dev) {
lock(&dev->lock);
  if (dev->resource != 5)
    dev->resource = 3;
unlock(&dev->lock);
}
\end{lstlisting}
\end{minipage}
\caption{Using locks to eliminate the data race}
\label{lock_example}
\end{figure}

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Figure~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.

### 2015-03-13 intro.tex
update in background section

In this paper, we present \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g. device drivers and filesystems). The key idea behind our approach is that a module can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques. We then show that by exploiting the race-freedom guarantees that our technique provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers. We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules from the Linux distribution.

>>>>

In this paper, we present \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux kernel modules (e.g. device drivers and filesystems). The key idea behind our approach is that a module can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers. We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.
